Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-15 analysis/paper/paper.Rmd
work on para of prior specification

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. Based on our hypothesis, we model a network with increased social inequality that is represented by low reciprocity, low transitivity, high popularity, high age homophily, and less proximity structure effects. We modeled four node attributes for homophily effects, including age, gender, ritual activity, and the degree of wealth, to explore the correlation between the tendency of connected nodes to form relationships and the network ties overall. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics. According to our hypothesis, the burial evidence from after the European arrival will have a better fit with our model compared with the burial evidence from before European contact.

@@

<!--Model fitting and prior setting--> 
After we set our model parameters, we simulated networks in a Bayesian framework using a Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithms allow estimation of posterior distributions through direct random sampling the posterior without assuming the prior comes from any specific distribution [@Hamra2013]. We can obtain a posterior distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on an algorithm. This enables a random or stochastic simulation in a long run where each move does not depend on the previous move. More chains ensures a more desirable posterior distribution that is close to the target distribution under study, or convergence. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and then switches to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations was `r burn_in` and the number of iterations after the burn-in was `r main_iter`. We set the number of iterations used to simulate a network y′ at each iteration to `r aux_iter`. We use the normal distribution for the priors, as is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017].

<!-- add text about anthropological assumptions of our prior values -->

@@

<!-- Goodness-of-fit assessment-->
The output from our two were models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of the model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, pre-European and post-European networks, with the distribution of our hypothesized model with higher popularity. We expect to see the network after the European presence has a better fit with the hypothesis, indicative of a structure with higher social differentiation.

>>>>

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. Based on our hypothesis, we model a network with increased social inequality that is represented by low transitivity and high popularity dyadic effects. We include nodal covariate effects for homophily, including age, gender, ritual activity, and the degree of wealth, to test whether nodes with similar attributes tend to form relationships presented by the network ties overall. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates showing the underlying mechanisms. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

<!--Model fitting--> 
After we set our model parameters, we simulated networks in a Bayesian framework using a Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithms allow estimation of posterior distributions through direct random sampling the posterior without assuming the prior comes from any specific distribution [@Hamra2013]. We can obtain a posterior distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on an algorithm. This enables a random or stochastic simulation in a long run where each move does not depend on the previous move. More chains ensures a more desirable posterior distribution that is close to the target distribution under study, or convergence. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and then switches to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations was `r burn_in` and the number of iterations after the burn-in was `r main_iter`. We set the number of iterations used to simulate a network y′ at each iteration to `r aux_iter`.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distribution for the priors is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the priors as low density for both network models. For covariate based on nodal attributes, such as age, gender, and wealth index, we set the priors to follow a normal distribution with mean 0 and standard deviation 5 (i.e. N(0, 5)) for both models. We set a negative covariate effect based on physical distance between burials (N(-1, 5)). To evaluate our anthropological assumption of increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for examining social inequality, especially for gwesp and gwedegree. For example, we set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on the ritual (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the prior for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to illustrate an increased social inequality. This prior information derives from the theory about the social structure of horizontal hierarchy, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate model at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for earlier network and N(-20, 6) for later network. We should note that the mean priors and standard deviation priors indicate relative differences for comparison.

<!-- Goodness-of-fit assessment-->
The output from our two  models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of the model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, pre-European and post-European networks, with the distribution of our hypothesized model with higher popularity. We expect to see the network after the European presence has a better fit with the hypothesis, indicative of a structure with higher social differentiation.
